A primer to numerical simulations: The perihelion motion of Mercury

### 2018-01-30 tex/MercuryNumerical.tex
Added python and vpython version number to section 4
Added footnote for prior vpython 7.0 import statement

In this section we present the numerical implementation as well as the visualization of planetary trajectories and in
particular the perihelion motion of Mercury.
We choose \python{}~\cite{Python} as programming language because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all necessary steps to create the simulation.
We provide online instructions for setting up \python{} and \vpython{} on different operating systems \cite{scripts}.
In \ref{sec:code}, we also provide a working example as well as possible extensions and template files online.

@@

So far we only fixed the length of the vectors. Next, we set up the initial directions, which will describe the motion in space.
We will build on the existing \python{} module \vpython{}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
With our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of Mercury is perpendicular to the vector which connects Mercury and Sun (see figure~\ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpython
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
We use the estimate of Eq.~(\ref{eq:check}) as a guidance to fix the time step $\Delta t$ as
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 20
\end{lstlisting}
Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following expressions\footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracies.
	Thus, the ordering and exact expressions for updating $a_{MS}$, $v_M$ and $r_{MS}$ are not unique. However, all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the definitions we show are in good balance between being numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
temp = 1 + alpha * rS / vec_rM_old.mag + beta * rL2 / vec_rM_old.mag**2
aMS  = c_a * temp / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \code{vector.mag} --- is an attribute of the vector and can be easily extracted.

>>>>

In this section we present the numerical implementation as well as the visualization of planetary trajectories and in
particular the perihelion motion of Mercury.
We choose \python{}~\cite{Python} as programming language because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all necessary steps to create the simulation.
In the following, we present code examples which tested for \python{} version \texttt{3.6.4} and \vpython{} version \texttt{7.3.2}.
We provide online instructions for setting up \python{} and \vpython{} on different operating systems \cite{scripts}.
In \ref{sec:code}, we also provide a working example as well as possible extensions and template files online.

@@

So far we only fixed the length of the vectors. Next, we set up the initial directions, which will describe the motion in space.
We will build on the existing \python{} module \vpython{}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
With our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of Mercury is perpendicular to the vector which connects Mercury and Sun (see figure~\ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpython
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}\footnote{Note that the statement \code{from vpython import vector} must be replaced by \code{from visual import vector} for \vpython{} versions prior \texttt{7.0}.}
We use the estimate of Eq.~(\ref{eq:check}) as a guidance to fix the time step $\Delta t$ as
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 20
\end{lstlisting}
Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following expressions\footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracies.
	Thus, the ordering and exact expressions for updating $a_{MS}$, $v_M$ and $r_{MS}$ are not unique. However, all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the definitions we show are in good balance between being numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
temp = 1 + alpha * rS / vec_rM_old.mag + beta * rL2 / vec_rM_old.mag**2
aMS  = c_a * temp / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \code{vector.mag} --- is an attribute of the vector and can be easily extracted.
